---
title: "Norwegian Basin Implementation"
subtitle: "StrathE2E2"
author: "Jack H Laverick<br> Douglas C Spiers<br> Michael R Heath"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  pagedreport::paged_hazelnuts:
    logo: "https://www.r-project.org/logo/Rlogo.svg"
    img_to_dark: TRUE
    logo_to_white: TRUE
    front_img: "img/Front.png"
    back_img: "img/Back.png"
    other_css: "tweaks.css"
knit: pagedown::chrome_print
main-color: "#000000"
secondary-color: "#737372"
main-font: "Quicksand"
bibliography: "IMP.bib"

---

```{r setup, include=FALSE}                

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "left")
rm(list=ls())
library(tidyverse)
library(sf)
library(gt)

source("@_Region file.R")
Period <- "2010-2019"

table_nums <- captioner::captioner(prefix = "Table ")

table.ref <- function(x) stringr::str_extract(table_nums(x), "[^:]*")

model <- StrathE2E2::e2e_read(implementation, Period, models.path="./Files") # Load implementation

Area <- read.csv(str_glue("Files/{implementation}/{Period}/Param/event_timing_{toupper(implementation)}_{Period}.csv")) %>% 
  filter(str_detect(Description, "km2")) %>% 
  .$Value                                                     # Get size of the model domain

```


# Introduction {-}                         

This document describes the configuration of StrathE2E2 for the Norwegian basin and its parameterisation to enable stationary state fitting for 2010-2070. These represent contrasting periods of environmental conditions. 

Volumetric and seabed habitat data define the physical configuration of the system. We regard these as being fixed in time. Similarly, we regard the physiological parameters of the ecology model as being fixed in time. Some of these are set from external data. The remainder are fitted, as detailed here. Changes in the model performance between the different time periods therefore stem from the hydrodynamic, hydro-chemical and fishery driving data. These are detailed in the ecological drivers and fishing fleet sections.

<br>

> Department of Mathematics and Statistics, University of Strathclyde, Glasgow, UK.
>
>E-mail: <m.heath@strath.ac.uk>
>
>The code written to support this parameterisation is available on [github](https://github.com/Jack-H-Laverick/MA.NorwegianBasin).

<br>

> **WARNING**: This is a working document, subject to update and revision.

# Model Domain {-}                         

The model splits the domain into three zones, inshore/shallow, offshore/shallow, and offshore/deep (Figure \@ref(fig:spatialschematic)). The inshore/shallow zone covers waters shallower than `r SDepth` m . The offshore zone covers the remaining area of the model domain (Figure \@ref(fig:habitats)). The offshore zone is divided further into a shallow and deep layer. The shallow layer represents water from the surface to `r SDepth` m depth, and shares a boundary with the inshore shallow zone. The offshore/deep zone covers the same area as the offshore/shallow zone, but represents water between `r SDepth` m and `r DDepth` m deep. There is a second internal boundary between the two offshore zones. 

![(#fig:spatialschematic) The spatial structure of StrathE2E; Ocean volumes and seafloor habitats. StrathE2E is built around a simplified spatial structure which represents shelf seas. These spatial units are connected to each other and to boundaries as shown to the right. The volumes connected to each spatial component are highlighted in blue.](img/schematic overhang.svg)

\newpage

The seafloor of the model domain is represented by 5 habitat types. There are three sediment classes for the inshore zone – fine (muddy, 1), medium (sandy, 2) and coarse (gravel, 3). The fourth class (rock, 0) represents an absence of soft sediment. The rock class has different geochemical properties and supports the kelp forests in the model food web. In the offshore zone, the seafloor being too deep to correctly be represented within StrathE2E, a imaginary floating layer of sediment called "overhang" will be added to the model to represent deepsea interactions (Figure \@ref(fig:habitats)). 

The perimeter of the model domain is defined by a `r Perimeter` m depth contour and the Norwegian coastline. Open ocean boundaries occur wherever there is no coastline. In consultation with local collaborators, we imposed additional boundaries to limit the eastward extent of the model domain at 42.2W and Southward along 33.75S, 54.35W to 36S, 50W. This coincides with changes in the pattern of fishing effort according to Global Fishing Watch, and Brazil's exclusive economic zone.

![(#fig:habitats) Map of the model domain. StrathE2E defines seabed sediment habitats as inshore (blues) or offshore (yellows). Within each zone, three sediment classes are represented – fine (muddy, 1), medium (sandy, 2) and coarse (gravel, 3). A fourth class (rock, 0) represents an absence of soft sediment. The overhang in the model domain which does not contact the sea floor is indicated in red. Sedimentary data are a composite of ___ and ____ [-@laverick21].](img/habitats.png)

# Fixed Physical {-}                       

## Background {-}                          

#### Water column inshore/shallow and offshore/deep zone area proportions and layer thicknesses; seabed habitat area proportions and sediment properties: {-}

Area proportions of depth zones and seabed habitats derived from 1/100th degree resolution atlas of seabed sediment properties [@laverick21]. The atlas provides gridded data sets of bathymetry, mean grain size, mud, sand and gravel content, porosity, permeability, organic nitrogen and carbon content, and natural disturbance by waves and bed shear stress.

#### Parameters for relationship between median grain size, sediment porosity and permeability. Permeability is used as the basis for estimating hydraulic conductivity which is a parameter in the representation of sediment processes in the model: {-}

Porosity (proportion by volume of interstitial water) and permeability of each sediment habitat were derived from median grain sizes using empirically-based relationships.

$$log_{10}(porosity) = p_3 + p_4\left(\frac{1}{1+e^{(\frac{-log_{10}(D_{50})-p_1}{p_2})}}\right)$$
D~50~ = median grain size (mm); parameters p~1~ = -1.227, p~2~ = -0.270, p~3~ = -0.436, p~4~ = 0.366 (Heath et al. 2015)

$$permeability = 10^{p_5}∙D_{50}^{*p_6}$$

where D~50~* = 0.11 ≤ D~50~ ≤ 0.50
p~5~ = -9.213, p~6~ = 4.615 [@heath15].

These relationships are coded into the StrathE2E2 R-package with the parameters in the csv setup file for the North Sea model. The parameters are probably a reasonable starting point for any future model of a new region. Derivation of the parameters is described in the following text sub-sections.

#### Parameters for in-built relationship between sediment mud content, and slowly degrading (refractory) organic nitrogen content of seabed sediments (see description in this document): {-}

Values for each sediment type derived from parameterised relationships between total organic nitrogen content of sediments (TON%, percent by weight), mud content (mud%, percent by weight) and median grain size (D~50~, mm).

$$mud\% = 10^{p_7}∙𝐷_{50}^{𝑝_8}$$
p~7~ = 0.657, p~8~ = -0.800

$$TON\% = 10^{𝑝_9}∙mud\%^{𝑝_{10}}$$

p~9~ = -1.965, p~10~ = 0.590

Proportion of TON estimated to be refractory = 0.9

These relationships are coded into the StrathE2E2 R-package with the parameters in the csv setup file for the North Sea model. The relationships and parameters are probably a reasonable starting point for any future model of a new region, though there are clear regional variations. Derivation of the parameters is described in the following sub-sections.

## Model area proportions {-}              

```{r area_proportions_cap}

tab_area_proportions_cap <- table_nums(name = "tab_area_proportions", 
    caption = stringr::str_glue("Area-proportions of the inshore and offshore zones and the thicknesses of the                                        water column layers. The sea surface area of the model domain is an estimated                                         {Area} km^2^."))
```
`r table_nums("tab_area_proportions")`
```{r area_proportions}

data_frame(Property = c("Sea-surface area proportion",
                        "Upper layer thickness (m)",
                        "Lower layer thickness (m)"),
           `Inshore/shallow` = c(model[["data"]][["physical.parameters"]][["x_shallowprop"]], 
                                 model[["data"]][["physical.parameters"]][["si_depth"]], 
                                 NA),
           `Offshore/deep` = c(1 - model[["data"]][["physical.parameters"]][["x_shallowprop"]], 
                               model[["data"]][["physical.parameters"]][["so_depth"]], 
                               model[["data"]][["physical.parameters"]][["d_depth"]])) %>% 
  gt() %>% 
  fmt_number(columns = c("Offshore/deep", "Inshore/shallow"), decimals = 4, use_seps = FALSE)


```

We derived the area-proportions of seabed habitat in the inshore and offshore zones from the atlas of seabed sediment properties from Laverick et al. [-@laverick21]. The atlas provides a range of seabed data for 1/100th degree cells over the Barents and Greenland Seas, including the presence of rock, the percentage of mud, sand and gravel fractions in the sediments, the whole–sediment mean grain size, and the natural disturbance rate by currents and waves. These values are derived from habitat classes used by the Norwegian Geological survey in partnership with the Russian Federal State Unitarian Research and Production Company for Geological Sea Survey (NGU-SEVMORGEO). We assigned the NGU-SEVMORGEO sediment classes as fine, medium, coarse, or absence of sediment habitats within each zone (Figure \@ref(fig:habitats)). The actual area of each habitat was then the sum of the areas of each set of assigned cells (`r table.ref("tab_habitat_proportions")`).

```{r habitat_proportions_cap}

tab_habitat_proportions_cap <- table_nums(name = "tab_habitat_proportions", 
    caption = stringr::str_glue("Area proportions and other characteristics of the 8 seabed habitat classes defined in the model by depth, rock or sediment type. The sea surface area of the model domain is an estimated {Area} km^2^. Grain size is the median in mm, Permeability in units of m^2^, nitrogen content in %dw."))
```
`r table_nums("tab_habitat_proportions")`
```{r habitat_proportions}

physical <- read.csv(str_glue("Files/{implementation}/{Period}/Param/physical_parameters_{toupper(implementation)}.csv"))
  
proportions <- which(str_detect(physical$Description, "Area_proportion")& physical$Value != 0)
porosity <- which(str_detect(physical$Description, "Defined_porosity"))
permeability <- which(str_detect(physical$Description, "Defined_permeability"))
nitrogen <- which(str_detect(physical$Description, "Defined_total_N"))
grain <- which(str_detect(physical$Description, "grain_size"))

data_frame(Habitat = c("S0", "S1", "S2", "S3", "D0"),
           Depth = c("Shallow","Shallow","Shallow","Shallow", "Deep"),
           Zone = c("Inshore","Inshore","Inshore","Inshore", "Offshore"),
           Sediment = c("None (Rock)", "Fine", "Medium", "Coarse","None (Overhang)"),
           `Area Proportion` = physical$Value[proportions],
           `Grain size` = c(NA, physical$Value[grain[1:3]], NA),
           Porosity = c(NA, physical$Value[porosity[1:3]], NA),
           Permeability = c(NA, physical$Value[permeability[1:3]], NA),
           `Nitrogen content` = c(NA, physical$Value[nitrogen[1:3]], NA)) %>% 
  group_by(Zone, Depth) %>% 
  gt() %>% 
  fmt_number(columns = c("Area Proportion", "Grain size", "Porosity", "Permeability", "Nitrogen content"), 
                         decimals = 4, use_seps = FALSE) %>% 
  cols_align(align = 'center',
             columns = vars(`Area Proportion`, `Grain size`, Porosity, Permeability, `Nitrogen content`)) %>% 
  tab_options(row_group.background.color = "whitesmoke")


```

## Sediment porosity {-}                   

Log-transformed porosity has been shown to have a sigmoidal relationship with log~10~(median grain size) (D~50~, mm) [@wilson18]:

$$log_{10}(porosity) = p_3 + p_4\left(\frac{1}{1+e^{(\frac{-log_{10}(D_{50})-p_1}{p_2})}}\right)$$

We use this relationship to calculate porosity for sea bed sediments in the Norwegian Sea (`r table.ref("tab_habitat_proportions")`), using an alternative parameterisation to Wilson [@pace21]. This alternative set of parameters extends the relationship to fine, muddy sediments (`r table.ref("tab_porosity")`). 

```{r porosity_cap}

tab_porosity_cap <- table_nums(name = "tab_porosity", 
    caption = "The four parameters for the function relating sediment porosity to median grain size. From Pace et al. (in review)")
```
`r table_nums("tab_porosity")`
```{r tab_porosity}

d50_to_pore <- which(str_detect(physical$Description, "porosity_and_grainsize"))

data_frame(Parameter = paste0("P", 1:4),
           `Fitted value` = physical$Value[d50_to_pore]) %>%
  pivot_wider(names_from = Parameter, values_from = `Fitted value`) %>% 
  gt()

```

## Hydraulic conductivity {-}              

Hydraulic conductivity (H, m.s^-1^) represents the ease with which fluids flow through the particle grain matrix. The related term ‘permeability’ (m^-2^) is a measure of the connectedness of the fluid filled void spaces between the particle grains. Permeability is a function only of the sediment matrix, whilst conductivity is a function of both the sediment and the permeating fluid, in particular the fluid viscosity and density.
Hydraulic conductivity is related to permeability by:

$$H = Permeability∙fluid\;density∙\frac{𝑔}{dynamic\;viscosity}$$

where: seawater density = 1027 kg.m^-3^ at salinity 35 and temperature 10°C; seawater dynamic viscosity = 1.48 x 10^-3^ kg.m^-1^.s^-1^ at salinity 35 and temperature 10°C; g = acceleration due to gravity = 9.8 m.s^-1^

Hence, $H = Permeability · 6.8004·10^6$ (m.s^-1^ at salinity 35 and temperature 10°C)

Whole sediment permeability can be related to the proportion of sediment classed as mud (D~50~ < 62 μm) [@pace21]. This relationship was used in the production of the Barents Sea sediment atlas [@laverick21], which provides the values used in this model implementation (`r table.ref("tab_habitat_proportions")`).

## Sediment organic nitrogen content {-}   

The magnitude of the static (refactory) organic nitrogen detritus pool in each sediment type is a required input to the model. The code includes an option to impute values from empirical relationships between total organic nitrogen (TON) and mud content, and between mud content and median grain size. This relationship has been documented in the North Sea implementation of the temperate StrathE2E2 package [@heath2021], and is based on sediment data off northeast Scotland.

To supplement the relationship within the package, predictions of sediment organic nitrogen content for the Norwegian Sea have been sourced from the sediment atlas [@laverick21] to parameterise StrathE2E directly (`r table.ref("tab_habitat_proportions")`).

# Fixed biological {-}                     

## Configuration parameters {-} 

### Assimilation efficiencies for each living guild in the model. {-}

Fixed parameters defining the proportion of ingested mass of food that contributes to new body tissue, after subtracting defecation and the metabolic costs of digestion and synthesis [@heath12].

### Biomass loss rates due to temperature-dependent metabolism for each living resource guild. {-}

Proportion of biomass lost to ammonia per day due to non-feeding related metabolism at a given reference temperature. Rates for individual guilds broadly related to typical body mass of representative species. Temperature dependency following a Q~10~ function.

### Q~10~ values for temperature dependent processes, and the Q~10~ reference temperature. {-}

Separate Q10 values for autotrophic uptake of nutrient, heterotrophic feeding, and heterotrophic metabolism based on literature data.

### Light intensity required to saturate autotrophic nutrient uptake. {-}

Light saturation intensity for nutrient uptake cannot be treated as a fitted value since it is confounded with other uptake parameters. Value estimated from survey of laboratory experiments.

### Annual weight specific fecundities of planktivorous and demersal fish guilds and the two benthos guilds in the model (suspension/deposit feeders and carnivore/scavenge feeders). {-}

Guild-level values derived by surveying the literature.

### Harvestable biomass density threshold for each resource guild. {-}

The living resource guilds in the model represent a mixture of harvestable and non-harvestable species, especially the invertebrate guilds. The density threshold parameter sets a limit for the guild biomass below which the harvestable species are assumed to be exhausted. Values set from analysis of trawl, plankton and benthos survey species biomass compositions.

### Minimum inedible biomass of carnivorous zooplankton. {-}

The carnivorous zooplankton guild is a key component of the food web, predated on by all the fish and top-predators. However it represents an extremely diverse range of fauna many of which are not edible in significant quantities by the guild predators, e.g. scyphomedusae. A minimum edible threshold is set to ensure that the guild as a whole cannot be extirpated by predation. The value is a rough estimate of scyphomedusae biomass.

## Event timing parameters (not fitted) {-}  

### Spawning start and end dates for fish and benthos {-}

For the fish guilds the dates were obtained from literature survey. The annual weight-specific fecundity is assumed to be shed uniformly between the start and end dates of spawning. 
Pelagic planktivorous fish in the model are assumed to be Norwegian Spring-Spawning herring (*Clupea harengus*)

### Recruitment start and end dates for fish and benthos {-}

Obtained from literature survey. The annual cohort of larvae/juveniles of each fish and benthos guild is assumed to recruit to the settled stage at a uniform daily rate between the start and end dates.

### Extra-domain stock biomass of migratory, and the proportion invading the domain each year. Start and end dates for the annual invasion, and start and end dates for the emigration. (see description below). {-}

The main migratory fish species undertaking a seasonal transit of the Norwegian Sea are the Atlantic mackerel and the Blue Whiting. Data on the North East Atlantic stock biomass, the proportion entering the Barents Sea and the timing of the migration, were derived from stock assessment literature [@ices2020]. 

## Event timing parameters (fitted) {-} 

Migratory fish in the Barents Sea model are assumed to be Atlantic mackerel and the Blue whiting. 
The spawning for Atlantic mackerel takes place off southwest Ireland in April and between February and April on the West of the British Isles for the Blue whiting. After spawning, both species rapidly migrate to summer feeding zones thousands of kilometers northwards along the continental shelf edge to the Norwegian and Barents Seas. [@iversen2002;@iversen2004;@payne2012]. Some spawning of Atlantic mackerel have been observed in the Norwegian Sea feeding area [@dossantosschmidt2024].

For the purposes of the model, we assume that there is no feedback between fishing and environmental conditions in the Norwegian Sea and the biomass and migration patterns of the whole northeast Atlantic mackerel or Blue whiting stock. In this version of StrathE2E2 the timing of immigration and emigration, and the mass influx across the ocean boundary during the annual immigration phase are treated as period-specific external driving data.

Data on the ‘global’ stock of northeast Atlantic mackerel and Blue whiting (wet biomass) are available from stock assessments [@ices2020], and converted to molar nitrogen mass using appropriate conversion ratios [@pedersen2021]. The proportion of the migrating stock entering the Norwegian Sea [@nottestad2016;@ekerhovd2010], and the timing of the inward and outward migrations [@iversen2002;@payne2012] are estimated from the literature. A residual proportion of the peak abundance in the North Sea remaining as residents (if any) is estimated from summer trawl survey data. The model setup code calculates the parameters which are needed in the ecology model.These are the only fixed (i.e. non-fitted) ecology model parameters which are period-specific.



```{r migrant_cap}

tab_migrant_cap <- table_nums(name = "tab_migrant", 
    caption = "Biological event timing parameters, constant across the 2011-2070 time period. The data are processed in the model setup to calculate the immigration flux parameters needed in the ecology model.")
```
`r table_nums("tab_migrant")`
```{r tab_migrant, fig.cap = tab_migrant_cap}

data_frame(Parameter = model[["data"]][["biological.events"]][["Description"]] %>% 
                        str_replace_all("_", " "),
           Value = model[["data"]][["biological.events"]][["Value"]]) %>% 
  .[1:13,] %>% 
  gt()

```
\newpage
`r table_nums("tab_migrant", display = "cite")` Continued.
```{r tab_migrantb}

data_frame(Parameter = model[["data"]][["biological.events"]][["Description"]] %>% 
                        str_replace_all("_", " "),
           Value = model[["data"]][["biological.events"]][["Value"]]) %>% 
  .[14:26,] %>% 
  gt()
```


# Ecological drivers {-}                   

Monthly resolution time-varying physical and chemical driving parameters for the model were derived from a variety of sources:

- Temperature, vertical mixing coefficients, volume fluxes, and boundary nutrient, detritus and phytoplankton concentrations from outputs of a NEMO-MEDUSA coupled hydro-geochemical model run at RCP85 with a 2005/2006 historical/future split [@yool13].
- Surface shortwave radiation, surface air temperature, and freshwater volume outflows from HadGEM2-ES model output [@jones11] used to force the NEMO-MEDUSA coupled hydro-geochemical model mentioned above [@yool13].
- River nitrate and ammonia concentrations taken from **TBD** by the [TBD](https://arcticgreatrivers.org/) [@holmes20].
- Atmospheric deposition of nitrate and ammonia from ISIMIP3a ([Inter-Sectoral Impact Model Intercomparison Project](https://data.isimip.org/10.48364/ISIMIP.759077.2); [@yang22])
- Oceanic Nitrate and ammonia data from **TBD**, published by **TBD** [@brand20; @brand20b; @brand20c; @brand20d].
- Remote sensing data products on Suspended Particulate Matter (Globcolour L3b; ftp://ftp.hermes.acri.fr/GLOB/merged/month/).
- Habitat disturbance due to tidal currents and waves from **TBD** [@laverick21].
- Wave height, period, and direction from the ERA-5 reanalysis monthly means accessed through [CDS](https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7?tab=overview) for 1980-2021 [@Hersbach19].

Details of how these data were processed are given below, supported by the [nemomedusR](https://jack-h-laverick.github.io/nemomedusR/) and [MiMeMo.tools](https://jack-h-laverick.github.io/MiMeMo.tools/) packages.

### Vertical mixing coefficients between the upper and lower layers of the deep zone: ### {-}

Vertical diffusivity from the NEMO-MEDUSA coupled hydro-geochemical model output [@yool13] was interpolated for each grid cell at the SDepth m boundary depth between the shallow and deep layers of the offshore zone. These values were summarised as monthly averages into period-specific climatological annual cycles of data for the 2010-2019 and 2040-2049 simulation periods.

### Monthly averaged temperatures for each water column layer: ### {-}

Derived by monthly averaging values at grid points within the inshore and vertical layers of the offshore zones from the NEMO-MEDUSA coupled hydro-geochemical model output [@yool13], weighted by grid point volumes. Values were summarised into period-specific climatological annual cycles of data for the 2010-2019 and 2040-2049 periods.

### Monthly averaged suspended particulate matter (SPM) concentrations (mg.m^-3^) in the shallow zone and the deep zone upper layer: ### {-}

Monthly averaged values of inorganic suspended particulate matter in sea water are available from the Globcolour project, starting from September 1997. These data are derived from satellite observations using the algorithm of Gohin [-@gohin11]. Data were downloaded from the ftp server (ftp://ftp.hermes.acri.fr/GLOB/merged/month/). We summarised these values as zonal statistics for the model domain to acquire a climatological annual cycle of data for the 2010-2019 simulation period only.

### Monthly average light attenuation coefficients for the inshore and offshore surface layers: ### {-}

Light attenuation in open water was parameterised from a linear relationship between the light attenuation coefficient and suspended particulate matter concentration (SPM) (Devlin et al., 2008).

### Monthly averaged daily integrated irradiance at the sea surface (E.m^-2^.d^-1^): ### {-}

Derived from HadGEM2-ES model output [@jones11] which forces the NEMO-MEDUSA model run used throughout our implementation. Monthly mean values were summarised into a climatological annual cycle of data for both the 2010-2019 and 2040-2049 periods.

### Monthly averaged daily atmospheric deposition rates of oxidised and reduced nitrogen onto the sea surface in the shallow and deep zones (mMN.m^-2^.d^-1^): ### {-}

Sourced from the "histsoc" files for a 1901 - 2021 hindcast as monthly averages [@yang22], available from [CDS](https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7?tab=overview). Monthly values were summarised into climatological annual cycles of monthly oxidised and reduced nitrogen deposition rates extracted for 2010-2019. Projections under different SSPs are also available to 2100.

### Monthly averaged, freshwater river inflow rates (expressed as a daily proportion of the receiving layer volume), and concentrations of oxidised and reduced dissolved inorganic nitrogen in the inflowing river waters (mMN.m^-3^): ### {-}

Freshwater inflow derived from HadGEM2-ES model output [@jones11] which forces the NEMO-MEDUSA model run used throughout our implementation. Monthly values were summaries into a climatological annual cycle of data for both the 2010-2019 and 2040-2049 periods.

The closest estimates of the concentrations of oxidised and reduced dissolved inorganic nitrogen in river water to the Barents Sea were from the river Ob', provided by the [Arctic Great Rivers Observatory](https://arcticgreatrivers.org/) [@holmes20]. We derived a climatological annual cycle of data for the 2011-2019 simulation period only.

### Volume fluxes into the model domain across open sea boundaries, and from the upper layer of the offshore/deep zone into the inshore/shallow zone, expressed as proportions of the receiving layer volume per day: ### {-}

Monthly averaged daily inflow and outflow volume fluxes derived by integrating daily mean velocities directed perpendicular to transects along the model domain boundary at grid points in each depth layer along transects through outputs from the NEMO-MEDUSA coupled hydro-geochemical model output [@yool13]. Monthly averaged daily inflow volume fluxes then divided by the volume of the receiving layer in the model domain to estimate a daily flushing rate. Period-specific climatological annual cycles of data used for 2010-2019 and 2040-2049 simulation periods.

### Mean concentrations of nitrate, ammonia, phytoplankton and suspended detritus (mMN.m^-3^), in adjacent ocean waters inflowing to the offshore/deep zone upper layer, adjacent ocean waters inflowing to the offshore/deep zone lower layer, and adjacent shelf waters inflowing to the inshore/shallow zone: ### {-}

NEMO-MEDUSA outputs included phytoplankton and suspended detritus, as well as Dissolved Inorganic Nitrogen (DIN). We calculated the depth-averaged concentrations for pixels within the shallow and deep layers of StrathE2E. We then sampled the pixels using the same transects around the model domain as for sampling volume fluxes. Only transects where water flowed into the model domain were sampled, and the average concentration of inflowing waters for target variables was calculated weighting by the flow rate across a transect and the cross-sectional area represented by a transect (average depth and length). Concentrations were then averaged into climatological annual cycles for both the 2010-2019 and 2040-2049 periods. 

DIN was decomposed into nitrate and ammonia concentrations using a ratio of ammonia:DIN derived from field observations collected during NERC Changing Arctic Ocean Cruises [@brand20; @brand20b; @brand20c; @brand20d]. Concentrations were averaged by depth layer into two correction factors across all samples located in the model domain and across all time steps.  

# Fishing fleet {-}                        

## Background {-} 

The key configuration data for the fishing fleet model are the definitions of the gears in terms of their power with respect to each of the harvestable resource guilds, discarding rates, processing-at-sea rates, and their seabed abrasion rates. These can be regarded as static parameters for each gear.

### Final processing {-}

1. Total catch by gear and guild was calculated by inflating landings according to EU discard rates before adding additional known discarded weight (described above for cetaceans, birds, planktivores and demersal fish).

2. Discarded weight was calculated as catch - landings.

3. Demersal non quota and quota limited were combined into a single Demersal guild for catch, landings and discards.

4. New discard rates reflecting all data sources were calculated as discarded weight / caught weight. When catch was 0 discard rates were set to 1 except for kelp harvesters which were assigned a discard rate of 0 for macrophytes.

5. Fishing power was calculated as catch / activity per gear.

# End {-}                                  
## Acknowledgements {-} 

Financial support for the development of the Brazilian Shelf implementation of StrathE2E2 came from the European Union's Horizon 2020 research and innovation programme (Mission Atlantic - No. 862428). We are grateful to:

- Andy Yool and the National Oceanography Centre for making outputs from the NEMO-MEDUSA model available to us, from which we extracted driving data for StrathE2E2.

## References {-} 



